Higher order unification for free!

### 2024-04-08 paper/main.tex
rework intro

Specifying and implementing a logic or a proof system from scratch requires
significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free is one uses the ML binders to represent the object
language ones. The second one that comes to mind is unification, a cornerstone
for proof construction and proof search. However reusing that brick may require
some additional work that we carry out in this paper.

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
In the notable of Higher Order Logic~\cite{books/sp/NipkowPW02},
the ML Isabelle is such a good fit that it implements an interactive proof
system for the object logic, and not just a specification.
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
We want to use the Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog: they back-chain lemmas taken
from a designated database of ``type class instances''.
For this reason we believe that $\lambda$Prolog, hence Elpi, is a good fit for
implementing such as form of automation. In this paper we focus on one aspect of
this work, namely \emph{how to reuse the unification of the ML Elpi in order to
implement the one used by the type-class solver}.

@@

>>>>

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
The use of these meta languages facilitated this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free. %, exploited in all works mentioned above.
The second one is that these meta languages come equipped with some form
of unification, a cornerstone in proof construction and proof search.

% exploited only
%in the notable case of of Higher Order Logic~\cite{books/sp/NipkowPW02}:
%the meta language Isabelle is such a good match for HOL that it could used to
%implement an interactive proof system for the object logic, in addition to
%it specification.\\
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
In particular we want to leverage the Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog that back-chain lemmas taken
from a designated database of ``type class instances'', hence
we can expect that Elpi is a good fit for
implementing such as form of automation. In this paper we focus on one aspect of
this work, namely \emph{how to reuse the higher order unification procedure
of the meta language in order to implement a type-class solver for the object
language}. As it turns out, re-using the unification of the meta language is
not a trivial task.

@@

\todo{cite isabelle's TC, that are baked in}
